A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-10-03 doc/paper1-submission/paper4.tex
Update results section slightly.

\documentclass[fleqn,12pt,lineno,english]{wlpeerj}
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
%\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
%\usepackage{xcolor}
%\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}

\makeatletter

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{algorithm,algpseudocode}
\usepackage{lmodern}
\definecolor{mylime}{RGB}{128,255,128}

@@

\begin{abstract}
We present a new supertree method that enables rapid estimation of a
summary tree on the scale of millions of leaves.  This supertree
summarizes a collection of input phylogenies and an input taxonomy.
We introduce formal requirements that such a supertree must satisfy in
order to transparently and justifiably represent the input trees.   In
addition to producing a tree, our method writes an annotations file
that reports which grouping in the input trees support each of the
groups in the supertree.

We compare the performance of this new method to a previous supertree
method on the inputs to the Open Tree of Life version 4, and find that
the number of displayed input splits increases from 35,518 to 39,639 and the
number of conflicting input splits decreases from 2760 to 1357.  The
new supertree method also improves on the previous supertree
construction method because it produces no unsupported branches, and
it avoids unnecessary polytomies.

@@

\section{Description of the propinquity pipeline}

@@

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input. In version 5.0 of the Open Tree
of Life synthetic tree, this version of the taxonomic tree contained
only 41,226 leaves (while the flag-cleaned version of OTT had 2,424,255
leaves).
\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}

@@

In version 5.0 of the summary tree, this decomposition procedure produced
5,545 subproblems, but only 1,586 of these were non-trivial to solve.
If a subproblem contains only two tips it is trivial; 2,443 subproblems
were trivial in this way. Similarly, if a subproblem contains only
2 trees it is trivial to solve because the solution will simply be
all of the groupings from the first tree combined with all of the
groupings from the second tree that are compatible with the first
tree; 2,838 subproblems were trivial in this way. The subproblem with
the largest number of tips contained 953 tips. The largest subproblem,
in terms of the number of input trees (including the taxonomic tree)
that were relevant, had 18 trees. Without decomposition, the supertree
problem would have had 512 input trees and 41,226 leaves.

@@

@@

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially. We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %

>>>>

\documentclass[fleqn,12pt,lineno,english]{wlpeerj}
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
%\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
%\usepackage{xcolor}
%\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\sloppy
\makeatletter

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\otl}{OTLv4}
\newcommand{\otlprop}{OTLv4/propinquity}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{algorithm,algpseudocode}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\definecolor{mylime}{RGB}{128,255,128}

@@

\begin{abstract}
We present a new supertree method that enables rapid estimation of a
summary tree on the scale of millions of leaves.  This supertree
method summarizes a collection of input phylogenies and an input taxonomy.
We introduce formal goals and criteria for such a supertree to satisfy
in order to transparently and justifiably represent the input
trees. In addition to producing a supertree, our method computes
annotations that describe which grouping in the input trees support
and conflict with each group in the supertree.

\medskip

We compare our supertree construction method to a previously published
supertree construction method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2760 to 1357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

\medskip

@@

\section{Description of the supertree method}

@@

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input.

\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}

@@

@@

\section{Results}

\subsection{Inputs}

We seek to assess the performance of our new supertree method by
comparing it to the supertree method of \cite{HinchliffEtAl2015}.  The
method of \cite{HinchliffEtAl2015} was used to construct the Open Tree
of Life v4 (\otl{}).  Therefore, in order to facilitate comparison, we 
applied our method to same input trees and taxonomy used by
\otl{}.  We used the same cleaning flags to trim the taxonomy and
input trees.  We refer to the resulting supertree as \otlprop{} since it was
constructed by applying the propinquity pipeline to the same inputs as
\otl{}.
The flag-cleaned version of OTT contained 2,424,255 leaves.  The
\otl{} supertree was constructed from 482 phylogenetic inputs,
containing a total of 45385 leaves, of which 41029 were unique.
After flag-cleaning and exemplification by propinquity, these trees contained 40323
unique tips.

\subsection{Subproblems}
In the \otlprob{} summary tree, the decomposition procedure produced
5,406 subproblems, but only 1,422 of these were non-trivial to solve.
If a subproblem contains only two tips it is trivial; 2,362 subproblems
were trivial in this way. Similarly, if a subproblem contains only
2 trees it is trivial to solve because the solution will simply be
all of the groupings from the first tree combined with all of the
groupings from the second tree that are compatible with the first
tree; 3,052 subproblems were trivial in this way. The subproblem with
the largest number of tips contained 946 tips. The largest subproblem,
in terms of the number of input trees (including the taxonomic tree)
that were relevant, had 16 trees. Without decomposition, the supertree
problem would have had 482 input trees and 41,226 leaves.

\subsection{splits}
We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %

@@

Here we have described the motivation and methodology used by the
propinquity tool that is currently used by the Open Tree of Life project
to build its summary tree. A modified version of the treemachine software
which built the summary tree described in the \citet{HinchliffEtAl2015}
paper is used by the project to serve the tree produced by propinquity
via Open Tree of Life APIs. The migration of summary tree construction
from treemachine (used for version 4) to propinquity (for all versions
from v5.0 to present) has decreased the computational time required
to construct a supertree from several hours to about 8 minutes (after
some preprocessing steps which only have to be performed when the
input taxonomy changes). The amount of RAM required during tree construction
has also decreased substantially.
